---
title: "Structure Learning Analysis"
author: "Jonah Yallop"
date: "2026-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(bnlearn)
library(Rgraphviz)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gtools)
```

# Load in networks

```{r}
alarm_fit <- read.bif("alarm.bif")
alarm_bn <- bn.net(alarm_fit)

andes_fit <- read.bif("andes.bif")
andes_bn <- bn.net(andes_fit)

water_fit <- read.bif("water.bif")
water_bn <- bn.net(water_fit)

insurance_fit <- read.bif("insurance.bif")
insurance_bn <- bn.net(insurance_fit)
```

# Define some standard parameters across tests

```{r}
sample_sizes = c(200, 500, 1000, 2000, 3000, 4000, 5000)
alphas = alphas <- c(0.001, 0.01, 0.05, 0.1)
scores = c("aic", "bic")
```

# 1. Run Sensitivity Testing

```{r, warning=FALSE}
# 1. Alpha sensitivity testing for PC and MMHC

# ALARM network

alpha_sensitivity_pc_alarm <- run_alpha_sensitivity_tests(
  network = alarm_fit,
  sample_sizes = sample_sizes,
  alphas = alphas,
  R = 25,
  test = "mi",
  max.sx = 3,
  seed_offset = 200,
  algorithm = "PC"
)

alpha_sensitivity_mmhc_alarm <- run_alpha_sensitivity_tests(
  network = alarm_fit,
  sample_sizes = sample_sizes,
  alphas = alphas,
  R = 25,
  test = "mi",
  max.sx = 3,
  seed_offset = 200,
  algorithm = "MMHC"
)

# INSURANCE network

alpha_sensitivity_pc_insurance <- run_alpha_sensitivity_tests(
  network = insurance_fit,
  sample_sizes = sample_sizes,
  alphas = alphas,
  R = 25,
  test = "mi",
  max.sx = 3,
  seed_offset = 200,
  algorithm = "PC"
)

alpha_sensitivity_mmhc_insurance <- run_alpha_sensitivity_tests(
  network = insurance_fit,
  sample_sizes = sample_sizes,
  alphas = alphas,
  R = 25,
  test = "mi",
  max.sx = 3,
  seed_offset = 200,
  algorithm = "MMHC"
)

# WATER network

alpha_sensitivity_pc_water <- run_alpha_sensitivity_tests(
  network = water_fit,
  sample_sizes = sample_sizes,
  alphas = alphas,
  R = 25,
  test = "mi",
  max.sx = 3,
  seed_offset = 200,
  algorithm = "PC"
)

alpha_sensitivity_mmhc_water <- run_alpha_sensitivity_tests(
  network = water_fit,
  sample_sizes = sample_sizes,
  alphas = alphas,
  R = 25,
  test = "mi",
  max.sx = 3,
  seed_offset = 200,
  algorithm = "MMHC"
)

# ANDES network

alpha_sensitivity_pc_andes <- run_alpha_sensitivity_tests(
  network = andes_fit,
  sample_sizes = sample_sizes,
  alphas = alphas,
  R = 3,
  test = "mi",
  max.sx = 3,
  seed_offset = 200,
  algorithm = "PC"
)

alpha_sensitivity_mmhc_andes <- run_alpha_sensitivity_tests(
  network = andes_fit,
  sample_sizes = sample_sizes,
  alphas = alphas,
  R = 10,
  test = "mi",
  max.sx = 3,
  seed_offset = 200,
  algorithm = "MMHC"
)
```


```{r, warning=FALSE}
# 2. Objective function sensitivity for HC and MMHC

# ALARM network

objective_sensitivity_hc_alarm <- run_objective_sensitivity_tests(
  network = alarm_fit,
  sample_sizes = sample_sizes,
  scores = scores,
  restart = 0,
  perturb = 1,
  algorithm = "HC"
)

objective_sensitivity_mmhc_alarm <- run_objective_sensitivity_tests(
  network = alarm_fit,
  sample_sizes = sample_sizes,
  scores = scores,
  restart = 0,
  perturb = 1,
  algorithm = "MMHC"
)

# INSURANCE network

objective_sensitivity_hc_insurance <- run_objective_sensitivity_tests(
  network = insurance_fit,
  sample_sizes = sample_sizes,
  scores = scores,
  restart = 0,
  perturb = 1,
  algorithm = "HC"
)

objective_sensitivity_mmhc_insurance <- run_objective_sensitivity_tests(
  network = insurance_fit,
  sample_sizes = sample_sizes,
  scores = scores,
  restart = 0,
  perturb = 1,
  algorithm = "MMHC"
)

# WATER network

objective_sensitivity_hc_water <- run_objective_sensitivity_tests(
  network = water_fit,
  sample_sizes = sample_sizes,
  scores = scores,
  restart = 0,
  perturb = 1,
  algorithm = "HC"
)

objective_sensitivity_mmhc_water <- run_objective_sensitivity_tests(
  network = water_fit,
  sample_sizes = sample_sizes,
  scores = scores,
  restart = 0,
  perturb = 1,
  algorithm = "MMHC"
)

# ANDES network

objective_sensitivity_hc_andes <- run_objective_sensitivity_tests(
  network = andes_fit,
  sample_sizes = sample_sizes,
  scores = scores,
  restart = 0,
  perturb = 1,
  algorithm = "HC"
)

objective_sensitivity_mmhc_andes <- run_objective_sensitivity_tests(
  network = andes_fit,
  sample_sizes = sample_sizes,
  scores = scores,
  restart = 0,
  perturb = 1,
  algorithm = "MMHC"
)
```

# Algorithm performance comparisons

Plan is to run 30 repeats for each network for each algorithm. Then average normalised SHD, precision and recall over the repeats.

```{r, warning=FALSE}
# MMHC performance

mmhc_performance_alarm <- run_mmhc_replicates(
  alarm_fit,
  sample_n = 5000,
  R = 30,
  score = "bic",
  test = "mi",
  alpha = 0.05
)

mmhc_performance_insurance <- run_mmhc_replicates(
  insurance_fit,
  sample_n = 5000,
  R = 30,
  score = "bic",
  test = "mi",
  alpha = 0.05
)

mmhc_performance_water <- run_mmhc_replicates(
  water_fit,
  sample_n = 5000,
  R = 30,
  score = "bic",
  test = "mi",
  alpha = 0.05
)

mmhc_performance_andes <- run_mmhc_replicates(
  andes_fit,
  sample_n = 5000,
  R = 30,
  score = "bic",
  test = "mi",
  alpha = 0.05
)

# HC performance

# ALARM
hc_performance_alarm <- run_hc_replicates(
  alarm_fit,
  sample_n = 5000,
  R = 30,
  score = "bic"
)

# INSURANCE
hc_performance_insurance <- run_hc_replicates(
  insurance_fit,
  sample_n = 5000,
  R = 30,
  score = "bic"
)

# WATER
hc_performance_water <- run_hc_replicates(
  water_fit,
  sample_n = 5000,
  R = 30,
  score = "bic"
)

# ANDES
hc_performance_andes <- run_hc_replicates(
  andes_fit,
  sample_n = 5000,
  R = 30,
  score = "bic"
)

# PC performance

# ALARM
pc_performance_alarm <- run_pc_replicates(
  alarm_fit,
  sample_n = 5000,
  R = 30,
  test = "mi",
  alpha = 0.05
)

# INSURANCE
pc_performance_insurance <- run_pc_replicates(
  insurance_fit,
  sample_n = 5000,
  R = 30,
  test = "mi",
  alpha = 0.05
)

# WATER
pc_performance_water <- run_pc_replicates(
  water_fit,
  sample_n = 5000,
  R = 30,
  test = "mi",
  alpha = 0.05
)

# ANDES
pc_performance_andes <- run_pc_replicates(
  andes_fit,
  sample_n = 5000,
  R = 30,
  test = "mi",
  alpha = 0.05
)
```

```{r}
# Combining results into algorithm level dataframes

combined_pc_performance <- bind_rows(
  pc_performance_alarm$summary %>% mutate(network = "ALARM"),
  pc_performance_insurance$summary %>% mutate(network = "INSURANCE"),
  pc_performance_water$summary %>% mutate(network = "WATER"),
  pc_performance_andes$summary %>% mutate(network = "ANDES")
)

combined_hc_performance <- bind_rows(
  hc_performance_alarm$summary %>% mutate(network = "ALARM"),
  hc_performance_insurance$summary %>% mutate(network = "INSURANCE"),
  hc_performance_water$summary %>% mutate(network = "WATER"),
  hc_performance_andes$summary %>% mutate(network = "ANDES")
)

combined_mmhc_performance <- bind_rows(
  mmhc_performance_alarm$summary %>% mutate(network = "ALARM"),
  mmhc_performance_insurance$summary %>% mutate(network = "INSURANCE"),
  mmhc_performance_water$summary %>% mutate(network = "WATER"),
  mmhc_performance_andes$summary %>% mutate(network = "ANDES")
)

combined_algorithms <- bind_rows(
  combined_mmhc_performance %>% mutate(algorithm = "MMHC"),
  combined_hc_performance %>% mutate(algorithm = "HC"),
  combined_pc_performance %>% mutate(algorithm = "PC")
)
```

```{r}
# Comparison values

mean_norm_shd_mmhc <- mean(combined_mmhc_performance$normalized_by_true)
mean_norm_shd_hc <- mean(combined_hc_performance$normalized_by_true)
mean_norm_shd_pc <- mean(combined_pc_performance$normalized_by_true)

mean_precision_mmhc <- mean(combined_mmhc_performance$precision)
mean_precision_hc <- mean(combined_hc_performance$precision)
mean_precision_pc <- mean(combined_pc_performance$precision)

mean_recall_mmhc <- mean(combined_mmhc_performance$recall)
mean_recall_hc <- mean(combined_hc_performance$recall)
mean_recall_pc <- mean(combined_pc_performance$recall)
```

```{r}
mean_norm_shd_hc
mean_norm_shd_pc
mean_norm_shd_mmhc

mean_precision_hc
mean_precision_pc
mean_precision_mmhc

mean_recall_hc
mean_recall_pc
mean_recall_mmhc
```

# Structural error decomposition

Splitting SHD values into structural and directional

```{r}
structural_decomp_df <- combined_algorithms %>%
  group_by(network, algorithm) %>%
  summarise(
    mean_norm_full_shd = mean(normalized_by_true, na.rm = TRUE),
    sd_shd = sd(normalized_by_true, na.rm = TRUE),
    mean_norm_skeleton_shd = mean(normalized_skeleton_shd, na.rm = TRUE),
    sd_skeleton_shd = sd(normalized_skeleton_shd, na.rm = TRUE),
    mean_norm_orientation_shd = mean(normalized_orientation, na.rm = TRUE),
    sd_orientation_shd = sd(normalized_orientation, na.rm = TRUE),
    .groups = "drop"
  )
```

# Edge recovery frequency analysis

```{r}
# Edge recovery across replicates
efreq_mmhc_alarm <- edge_recovery_frequency(alarm_fit, mmhc_performance_alarm$raw)
names(efreq_mmhc_alarm) <- c("edge", "mmhc_freq", "mmhc_prop")

efreq_hc_alarm <- edge_recovery_frequency(alarm_fit, hc_performance_alarm$raw)
names(efreq_hc_alarm) <- c("edge", "hc_freq", "hc_prop")

efreq_pc_alarm <- edge_recovery_frequency(alarm_fit, pc_performance_alarm$raw)
names(efreq_pc_alarm) <- c("edge", "pc_freq", "pc_prop")

most_recovered_edges_mmhc <- efreq_mmhc_alarm[order(-efreq_mmhc_alarm$prop)[1:10], ] # top 10 most often recovered
least_recovered_edges_mmhc <- efreq_mmhc_alarm[order(efreq_mmhc_alarm$prop)[1:10], ] # top 10 most often recovered

most_recovered_edges_hc <- efreq_hc_alarm[order(-efreq_hc_alarm$prop)[1:10], ] # top 10 most often recovered
least_recovered_edges_hc <- efreq_hc_alarm[order(efreq_hc_alarm$prop)[1:10], ] # top 10 most often recovered
```

```{r}
# Combine results from all algorithms
combined_edge_frequencies <- efreq_mmhc_alarm %>%
  left_join(efreq_pc_alarm, by = c("edge")) %>%
  left_join(efreq_hc_alarm, by = c("edge"))

combined_edge_frequencies <- bind_rows(
  efreq_pc_alarm %>% mutate(algo = "PC"),
  efreq_hc_alarm %>% mutate(algo = "HC"),
  efreq_mmhc_alarm %>% mutate(algo = "MMHC")
)

combined_edge_frequencies <- combined_edge_frequencies %>% 
  arrange(desc(hc_freq), desc(mmhc_freq))

edge_bands <- combined_edge_frequencies %>%
  mutate(
    pc_band   = assign_band(pc_prop),
    hc_band   = assign_band(hc_prop),
    mmhc_band = assign_band(mmhc_prop)
  )

pc_edge_summary <- edge_bands %>%
  count(pc_band) %>%
  mutate(prop = n / sum(n))

hc_edge_summary <- edge_bands %>%
  count(hc_band) %>%
  mutate(prop = n / sum(n))

mmhc_edge_summary <- edge_bands %>%
  count(mmhc_band) %>%
  mutate(prop = n / sum(n))

edge_summary_table <- full_join(pc_edge_summary, hc_edge_summary, 
                           by = c("pc_band" = "hc_band")) %>%
  full_join(mmhc_edge_summary, by = c("pc_band" = "mmhc_band")) %>%
  rename(
    band = pc_band,
    pc_prop = prop.x,
    hc_prop = prop.y,
    mmhc_prop = prop
  )

# Keep top edges (e.g. by average frequency)
top_edges <- df %>%
  group_by(edge) %>%
  summarise(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop)) %>%
  slice(1:12) %>%
  pull(edge)

df_plot <- df %>% filter(edge %in% top_edges)

ggplot(df_plot, aes(x = reorder(edge, -prop), y = prop, fill = algo)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Edge", y = "Recovery Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Reverse engineering edge decisions

```{r, warning=FALSE}
res <- evaluate_pc_once(alarm_fit, data = rbn(alarm_fit, n = 5000), alpha = 0.05)

pc_fit <- res$pc_fit
```

```{r}
res_sep <- find_sepset(rbn(alarm_fit, n = 1000), "EXPCO2", "VENTLUNG")

res_sep
```

## Runtime analysis

```{r}
hc_water_runtime <- mean(hc_performance_water$summary$runtime_sec)
hc_insurance_runtime <- mean(hc_performance_insurance$summary$runtime_sec)
hc_alarm_runtime <- mean(hc_performance_alarm$summary$runtime_sec)
hc_andes_runtime <- mean(hc_performance_andes$summary$runtime_sec)

pc_water_runtime <- mean(pc_performance_water$summary$runtime_sec)
pc_insurance_runtime <- mean(pc_performance_insurance$summary$runtime_sec)
pc_alarm_runtime <- mean(pc_performance_alarm$summary$runtime_sec)
pc_andes_runtime <- mean(pc_performance_andes$summary$runtime_sec)

mmhc_water_runtime <- mean(mmhc_performance_water$summary$runtime_sec)
mmhc_insurance_runtime <- mean(mmhc_performance_insurance$summary$runtime_sec)
mmhc_alarm_runtime <- mean(mmhc_performance_alarm$summary$runtime_sec)
mmhc_andes_runtime <- mean(mmhc_performance_andes$summary$runtime_sec)

runtime_table <- data.frame(
  Network = c("ALARM", "INSURANCE", "WATER", "ANDES"),
  PC      = c(pc_alarm_runtime, pc_insurance_runtime, pc_water_runtime, pc_andes_runtime),
  HC      = c(hc_alarm_runtime, hc_insurance_runtime, hc_water_runtime, hc_andes_runtime),
  MMHC    = c(mmhc_alarm_runtime, mmhc_insurance_runtime, mmhc_water_runtime, mmhc_andes_runtime)
)

runtime_table
```

