---
title: "Football_BN_v4"
author: "Jonah Yallop"
date: "2025-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load required packages

```{r, include=FALSE}
library(bnlearn)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(purrr)
library(gRain)
library(igraph)
library(verification)
library(zoo)
library(data.table)
library(glue)
```

# 2. Load required data
- Historical match results from 1993
-- Used to build up the CPT for rank -> match result
- Current season fixtures
-- Used for predictions
- Sesaon end tables since 1993
-- Used for historical rank calculation
- Current table
-- Used to get current expected points and projected goal difference
- ELO ratings
-- Used for fitting the CPT

# 2.1. Read in raw data

```{r}
# Load in all matches since 1993/94 to build up conditional probability table
historical_matches <- read.csv("Data/matches_processed.csv")

# Load in 25/26 fixtures for prediction
fixtures_2526 <- read.csv("Data/2025_26_fixtures.csv")

# Load in 25/26 results for prediction
results_2526 <- read.csv("Data/2025_26_results.csv")

# Load in tables from 1993/94 - 2023/24
season_tables <- read.csv("Data/epl_tables.csv")

# Current premier league table
current_table_gw7 <- read.csv("Data/current_table.csv")
current_table_gw26 <- read.csv("Data/current_table_gw26.csv")

current_table <- current_table_gw7

# Load in current ELO ratings
elo_ratings <- read.csv("Data/elo_ratings.csv")

# Load in historical ELO ratings
historical_elo <- read.csv("Data/historical_elo.csv")
```

# 2.2. Initial data processing

```{r, warning=FALSE}
# Add season column to current fixtures
fixtures_2526['season'] <- "2025/26"

season_tables <- season_tables %>%
  rename(season = season_end_year) %>%
  mutate(season = season - 1)

# Filter for seasons since 2000 - due to ELO availability
season_tables <- season_tables[season_tables$season >= 2000,]

season_tables <- subset(season_tables, select = - c(X))

# Convert season column to year integer
historical_matches <- historical_matches %>%
  mutate(season = str_sub(season, 1, -4)) %>%
  mutate(season = as.integer(season))

# Filter for matches since 2000 - due to ELO availability
historical_matches <- historical_matches[historical_matches$season >= 2000,]

names(historical_elo)[names(historical_elo) == 'date'] <- 'elo_date'

# ----------------------------------------------
# Merging start of season ELO values onto tables
# ----------------------------------------------

# Add date column for merge
season_tables <- season_tables %>%
  mutate(elo_date = paste0("01/08/", season))

# Merge on ELO for each team at the start of the season
season_tables <- season_tables %>%
  left_join(historical_elo %>% dplyr::select(team, elo_date, elo),
            by = c("elo_date", "team"))

season_tables <- season_tables[!is.na(season_tables$elo),]

# -----------------------------
# Merging ELO onto each fixture
# -----------------------------

names(historical_elo)[names(historical_elo) == 'elo_date'] <- 'date'

historical_elo <- historical_elo[historical_elo$country == "ENG",]

# Convert dates
setDT(historical_matches)
setDT(historical_elo)

historical_matches[, date := as.IDate(strptime(date, "%d/%m/%Y"))]
historical_elo[, date := as.IDate(strptime(date, "%d/%m/%Y"))]

# Set key on ELO table
setorder(historical_elo, team, date)
setkey(historical_elo, team, date)

# ------------------------------------
# Assign ELO onto matches (home teams)
# ------------------------------------

historical_matches[, home_elo := historical_elo[.SD, on = .(team = home_team, date), elo, roll = TRUE]]

# Compute first ELO date per team and force NA for earlier matches
min_elo_by_team <- historical_elo[, .(min_elo_home = min(date)), by = team]

# Merge min ELO on
historical_matches <- merge(
  historical_matches,
  min_elo_by_team,
  by.x = "home_team",
  by.y = "team",
  all.x = TRUE,
  sort = FALSE
)

# ------------------------------------
# Assign ELO onto matches (away teams)
# ------------------------------------

historical_matches[, away_elo := historical_elo[.SD, on = .(team = away_team, date), elo, roll = TRUE]]

min_elo_by_team <- historical_elo[, .(min_elo_away = min(date)), by = team]

historical_matches <- merge(
  historical_matches,
  min_elo_by_team,
  by.x = "away_team",
  by.y = "team",
  all.x = TRUE,
  sort = FALSE
)

# Force values to be NA if earlier than data source
historical_matches[date < min_elo_home, home_elo := NA]
historical_matches[date < min_elo_away, away_elo := NA]
historical_matches[, min_elo_home := NULL]
historical_matches[, min_elo_away := NULL]

# Reorder columns
order <-  c("match_id", "date", "season", "home_team", "away_team", "home_goals", "away_goals", "result", "home_elo", "away_elo")

# Changing the structure of results
results_2526 <- results_2526 %>%
  rename(
    match_number = Match.Number,
    gameweek     = Round.Number,
    home_team    = Home.Team,
    away_team    = Away.Team,
    result       = Result,
    location     = Location,
    date         = Date
  )

results_2526 <- results_2526 %>%
  separate(result, into = c("home_goals", "away_goals"), sep = "-", convert = TRUE) %>%
  mutate(
    result = case_when(
      home_goals > away_goals ~ "H",
      home_goals < away_goals ~ "A",
      TRUE ~ "D"
    )
  )
```

# Helper functions

```{r}
# Function to map team names to the same standard

# Name mapping, current name = new name
team_map <- c(
  "Man Utd" = "Manchester Utd",
  "Man City" = "Manchester City",
  "Spurs" = "Tottenham",
  "Nott'm Forest" = "Nottingham Forest",
  "Newcastle" = "Newcastle Utd",
  "Leeds" = "Leeds United"
)

align_team_names <- function(table, mapping_dict, cols = c("home_team", "away_team")) {
  table <- table %>%
    mutate(
      across(
        all_of(cols),
        ~ recode(.x, !!!mapping_dict, .default = .x)
      )
    )
  
  return(table)
}

results_2526 <- align_team_names(results_2526, team_map)
```


```{r}
# -----------------------
# 1. Function to map end of season points to 1-14 rank scale
# -----------------------

points_to_rank <- function(points) {
  cuts <- c(-Inf, 24, 29, 34, 39, 44, 49, 54, 59, 64, 69, 74, 79, 84, Inf)
  # label 14..1 so higher points -> 1 (strongest)
  rank_num <- cut(points, breaks = cuts, labels = 14:1, right = TRUE)
  as.integer(as.character(rank_num))
}

# -----------------------
# 2. Function to compute expected end of season points incorporating prev 5 seasons
# As was done in pi-football
# -----------------------

compute_component1_mean <- function(season_points, current_season_points, cur_season) {
  
  # --- Teams to process
  teams <- unique(current_season_points$team)
  
  # --- Prior
  prior_mean <- 57
  prior_var  <- 300
  
  # --- Variance inflation for previous seasons
  prev_add <- c(3^6, 3^5, 3^4, 3^3, 3^2)   # oldest -> most recent
  names(prev_add) <- paste0("prev", 5:1)
  
  # --- Process each team
  out <- lapply(teams, function(team_name) {
    
    # --- Historical points (length guaranteed to be 5)
    prev_pts <- sapply(1:5, function(i) {
      s_i <- cur_season - i
      v <- season_points %>%
           filter(season == s_i, team == team_name) %>%
           pull(points)
      if (length(v) == 0) {
        NA_real_
      } else {
        as.numeric(v[1])  # ensure a numeric scalar
      }
    })

    # --- Current season points
    cur_row <- current_season_points %>% filter(team == team_name)
    cur_points <- ifelse(nrow(cur_row) == 0, NA_real_, cur_row$total_points)
    matches_played <- ifelse(nrow(cur_row) == 0, NA_integer_, cur_row$matches_played)
    
    # --- Combine inputs
    inputs <- c(prev_pts, current = cur_points)
  
    # --- Variances
    prev_vars <- prior_var + prev_add
    cur_var <- if (is.na(matches_played)) {
      prior_var + mean(prev_add)
    } else {
      exponent <- 6 - ((matches_played / 38) * 4)
      prior_var + 3^exponent
    }
    vars <- c(prev_vars, current = cur_var)
    
    # --- Observed indices
    obs_idx <- which(!is.na(inputs))
    
    # --- Posterior mean and variance
    if (length(obs_idx) == 0) {
      mean_post <- NA_real_
      var_post  <- NA_real_
    } else {
      precisions <- 1 / vars[obs_idx]
      weighted_sum <- sum(inputs[obs_idx] * precisions)
      post_precision <- sum(precisions)
      
      mean_post <- (prior_mean / prior_var + weighted_sum) / (1 / prior_var + post_precision)
      var_post  <- 1 / (1 / prior_var + post_precision)
    }
    
    tibble(team = team_name, Mean = mean_post, Var = var_post)
  })
  
  bind_rows(out)
}

# -----------------------
# 3. Function to discretise goal difference and ELO
# -----------------------

discretise_features <- function(df,
                                goaldiff_breaks = c(-Inf, -30, -20, -10, 0, 10, 20, 30, Inf),
                                goaldiff_labels = NULL,
                                elo_breaks = NULL,   # if NULL will use quantiles
                                prev_breaks = NULL,  # if NULL will use quantiles
                                elo_labels = NULL,
                                prev_labels = NULL) {
  # GoalDiff labels: if not supplied, create labels that match number of intervals
  n_gd_intervals <- length(goaldiff_breaks) - 1
  if (is.null(goaldiff_labels)) {
    lefts  <- head(goaldiff_breaks, -1)
    rights <- tail(goaldiff_breaks, -1)
    goaldiff_labels <- vapply(seq_len(n_gd_intervals),
                              function(i) paste0("(", lefts[i], ",", rights[i], "]"),
                              character(1), USE.NAMES = FALSE)
  } else {
    if (length(goaldiff_labels) != n_gd_intervals) {
      stop("goaldiff_labels length must equal number of intervals (length(goaldiff_breaks)-1).")
    }
  }

  df$goal_difference_bin <- cut(df$projected_goal_difference,
                                breaks = goaldiff_breaks,
                                labels = goaldiff_labels,
                                include.lowest = TRUE)

  # ELO
  if ("elo" %in% colnames(df)) {
    if (is.null(elo_breaks)) {
      elo_breaks <- unique(quantile(df$elo, probs = c(0, .25, .5, .75, 1), na.rm = TRUE))
  }
    if (is.null(elo_labels)) {
      n_elo_intervals <- length(elo_breaks) - 1
      elo_labels <- paste0("rating", seq_len(n_elo_intervals))
  } else {
    if (length(elo_labels) != (length(elo_breaks) - 1)) {
      stop("elo_labels length must equal number of intervals (length(elo_breaks)-1).")
    }
  }
  df$elo_bin <- cut(df$elo, breaks = elo_breaks, labels = elo_labels, include.lowest = TRUE)
  
  } else {
      if (is.null(elo_breaks)) {
        elo_breaks <- unique(quantile(df$rating, probs = c(0, .25, .5, .75, 1), na.rm = TRUE))
    }
    if (is.null(elo_labels)) {
      n_elo_intervals <- length(elo_breaks) - 1
      elo_labels <- paste0("rating", seq_len(n_elo_intervals))
    } else {
      if (length(elo_labels) != (length(elo_breaks) - 1)) {
        stop("elo_labels length must equal number of intervals (length(elo_breaks)-1).")
    }
  }
  df$elo_bin <- cut(df$rating, breaks = elo_breaks, labels = elo_labels, include.lowest = TRUE)
    
  }

  return(df)
}

# -----------------------
# 3. Function to discretise goal difference
# -----------------------

# Add goal difference bin to season tables
discretise_gd <- function(df,
                                goaldiff_breaks = c(-Inf, -30, -20, -10, 0, 10, 20, 30, Inf),
                                goaldiff_labels = NULL) {
  
  # GoalDiff labels: if not supplied, create labels that match number of intervals
  n_gd_intervals <- length(goaldiff_breaks) - 1
  if (is.null(goaldiff_labels)) {
    lefts  <- head(goaldiff_breaks, -1)
    rights <- tail(goaldiff_breaks, -1)
    goaldiff_labels <- vapply(seq_len(n_gd_intervals),
                              function(i) paste0("(", lefts[i], ",", rights[i], "]"),
                              character(1), USE.NAMES = FALSE)
  } else {
    if (length(goaldiff_labels) != n_gd_intervals) {
      stop("goaldiff_labels length must equal number of intervals (length(goaldiff_breaks)-1).")
    }
  }

  df$goal_difference_bin <- cut(df$gd,
                                breaks = goaldiff_breaks,
                                labels = goaldiff_labels,
                                include.lowest = TRUE)

  return(df) 
}

# -----------------------
# 4. Discrete normal builder
# -----------------------

# Discrete-normal builder
discrete_normal_probs <- function(mu, sd, n) {
  x <- seq_len(n)
  p <- dnorm(x, mean = mu, sd = sd)
  if (all(p == 0)) p <- rep(1/n, n)
  p / sum(p)
}

# -----------------------
# 5. Building CPT table
# -----------------------

# Build CPT by shifting base marginal distribution
build_cpt_from_base_shift3 <- function(base_rank_p,
                                       parent_grid,
                                       shift_currexp = NULL,
                                       shift_gd = NULL,
                                       shift_elo = NULL,
                                       sd_multiplier = 1.0,
                                       alpha = 1.0,
                                       pseudo_scale = 1000) {
  # base_rank_p: numeric vector length n_rank summing to 1
  # parent_grid: data.frame with columns current_exp_points_bin, goal_difference_bin, elo_bin
  # shift_*: named numeric vectors, names must match factor levels in parent_grid
  n_rank <- length(base_rank_p)
  ranks <- seq_len(n_rank)
  mu0 <- sum(ranks * base_rank_p)                       # baseline mean rank
  sd0  <- sqrt(sum((ranks - mu0)^2 * base_rank_p))     # baseline sd
  if (sd0 == 0) sd0 <- 1.0

  n_configs <- nrow(parent_grid)
  cpt_mat <- matrix(NA, nrow = n_rank, ncol = n_configs)
  colnames(cpt_mat) <- apply(parent_grid, 1, paste, collapse = "|")

  for (i in seq_len(n_configs)) {
    row <- parent_grid[i, , drop = TRUE]
    s_curr <- if (!is.null(shift_currexp)) shift_currexp[[ as.character(row$current_exp_points_bin) ]] else 0
    s_gd   <- if (!is.null(shift_gd))         shift_gd[[ as.character(row$goal_difference_bin) ]] else 0
    s_elo  <- if (!is.null(shift_elo))        shift_elo[[ as.character(row$elo_bin) ]] else 0
    total_shift <- sum(c(s_curr, s_gd, s_elo), na.rm = TRUE)
    mu_new <- mu0 - total_shift    # decrease numeric rank for positive shift
    sd_new <- sd0 * sd_multiplier
    p_new <- discrete_normal_probs(mu_new, sd_new, n_rank)

    # Dirichlet-like smoothing: convert p_new to pseudo-counts, add alpha, renormalize
    if (!is.null(alpha) && alpha > 0) {
      counts <- p_new * pseudo_scale
      p_smoothed <- (counts + alpha) / (sum(counts) + alpha * n_rank)
      p_new <- p_smoothed
    }
    cpt_mat[, i] <- p_new
  }
  list(cpt_mat = cpt_mat, parent_labels = colnames(cpt_mat))
}

# -----------------------
# 6. Converting CPT matrix to array for bnlearn
# -----------------------

# Convert cpt matrix to bnlearn array (rank x currExp x GD x Elo)
# Corrected matrix -> cpt array conversion
matrix_to_cpt_array3 <- function(cpt_mat, parent_levels, rank_names = NULL) {
  # cpt_mat: matrix with rows = ranks, cols = parent configurations
  # parent_levels: named list in the order: current_exp_points_bin, goal_difference_bin, elo_bin
  if (is.null(rank_names)) rank_names <- as.character(seq_len(nrow(cpt_mat)))
  # dims: n_rank, nCurrExp, nGD, nElo
  dims <- c(nrow(cpt_mat), vapply(parent_levels, length, integer(1)))
  dimnames_list <- c(list(rank = rank_names), parent_levels)
  arr <- array(NA_real_, dim = dims, dimnames = dimnames_list)

  # Build grid in the same order as parent_levels (expand.grid uses order of list)
  grid <- expand.grid(parent_levels, stringsAsFactors = FALSE)

  # Create column ids from grid and compare to column names of cpt_mat
  col_ids_grid <- apply(grid, 1, paste, collapse = "|")
  if (!is.null(colnames(cpt_mat))) {
    if (!all(col_ids_grid == colnames(cpt_mat))) {
      stop("Column ordering/labels of parent_grid (from expand.grid) do not match colnames(cpt_mat).\n",
           "First few grid ids: ", paste(head(col_ids_grid, 6), collapse = ", "), "\n",
           "First few cpt_mat colnames: ", paste(head(colnames(cpt_mat), 6), collapse = ", "))
    }
  }

  # For every column j place cpt_mat[,j] into the corresponding slice
  for (j in seq_len(ncol(cpt_mat))) {
    # pick the parent values for this column
    parent_vals <- as.list(grid[j, , drop = TRUE])
    # find position indices for each parent level
    indices <- lapply(seq_along(parent_vals), function(k) {
      match(as.character(parent_vals[[k]]), dimnames_list[[k + 1]])
    })
    # now assign into arr: arr[, idx2, idx3, idx4] <- cpt_mat[, j]
    # build the call dynamically to support N parents (here 3 parents expected)
    if (length(indices) == 3) {
      arr[, indices[[1]], indices[[2]], indices[[3]]] <- cpt_mat[, j]
    } else {
      # generic multi-parent assignment (supports different number of parent dims)
      # build index list: first is full rank slice, then each parent index
      idx_list <- c(list(seq_len(nrow(cpt_mat))), indices)
      # use do.call with '[' replacement via an intermediate approach
      # convert arr into vector and compute linear indices is more complex; simplest: use recursive assignment
      # fallback for unusual parent counts:
      stop("matrix_to_cpt_array3 currently supports exactly 3 parent dimensions in parent_levels.")
    }
  }
  return(arr)
}

# -----------------------------------------------------
# 7. Function to calculate a conditional probability table
# -----------------------------------------------------

# Building a conditional probability table giving the probability of a rank x
# team beating a rank y team

build_empirical_cpt <- function(historical_matches, season_points, lambda=1) {
  # 1. Prepare season points and ranks
  sp <- season_points %>%
    mutate(rank = points_to_rank(points))
  
  # 2. Merge ranks for home and away teams
  HM <- historical_matches %>%
    left_join(sp %>% dplyr::select(season, team, rank),
              by = c("season", "home_team"="team")) %>%
    rename(rank_home = rank) %>%
    left_join(sp %>% dplyr::select(season, team, rank),
              by = c("season", "away_team"="team")) %>%
    rename(rank_away = rank) %>%
    mutate(rank_home = if_else(is.na(rank_home), 12, rank_home)) %>%
    mutate(rank_away = if_else(is.na(rank_away), 12, rank_away))
  
  # 3. Compute match results
  HM <- HM %>% 
    mutate(result = case_when(
      home_goals > away_goals ~ "H",
      home_goals == away_goals ~ "D",
      home_goals < away_goals  ~ "A"
    ))
  
  # 4. Count outcomes for each (rank_home, rank_away, result)
  cpt <- HM %>%
    group_by(rank_home, rank_away, result) %>%
    summarise(n = n(), .groups = "drop") %>%
    tidyr::complete(rank_home = 1:14, rank_away = 1:14, result = c("H","D","A"),
                    fill = list(n = 0)) %>%
    group_by(rank_home, rank_away) %>%
    # Apply Dirichlet smoothing: add lambda pseudo-count to each outcome
    mutate(total = sum(n),
           p = (n + lambda) / (total + 3 * lambda)) %>%
    ungroup()
  
  # 5. Pivot to wide format
  cpt_wide <- cpt %>%
    dplyr::select(rank_home, rank_away, result, p) %>%
    tidyr::pivot_wider(
      names_from = result,
      values_from = p, 
      names_prefix = "p_"
    ) %>%
    arrange(rank_home, rank_away)
  
  # 6. Build 3D array for bnlearn CPT
  res_array <- array(0, dim = c(14, 14, 3),
                     dimnames = list(rank_home = as.character(1:14),
                                     rank_away = as.character(1:14),
                                     result = c("H","D","A")))
  
  for(i in seq_len(nrow(cpt_wide))) {
    rh <- as.character(cpt_wide$rank_home[i])
    ra <- as.character(cpt_wide$rank_away[i])
    res_array[rh, ra, "H"] <- cpt_wide$p_H[i]
    res_array[rh, ra, "D"] <- cpt_wide$p_D[i]
    res_array[rh, ra, "A"] <- cpt_wide$p_A[i]
  }

  list(cpt_wide = cpt_wide, cpt_array = res_array, ranked_matches = HM)
}

# ------------------
# 8. If else helper
# ------------------

`%||%` <- function(a,b) if (!is.null(a) && !any(is.na(a))) a else b
```

# 4. Data manipulation and node calculation

```{r}
# Calculate mean and variance for each team
# Mean of current projected points and previous 5 seasons
# 2025/26 season
current_mean_variance_df <- compute_component1_mean(
  season_points = season_tables,
  current_season_points = current_table,
  cur_season = 2026 # 2025 would be used to predict 24/25 season
)

# Merge ELO ratings onto current table
current_table <- merge(current_table, elo_ratings, by = "team")

# Project goal difference forward to the end of the season
current_table$projected_goal_difference <- current_table$goal_difference / current_table$matches_played * 38

# Discretise ELO and goal difference into bins
current_table_discrete <- discretise_features(
  df = current_table,
)

# Merging on the expected points generated according to pi football - incorporating current + prev 5 seasons
current_table_discrete <- merge(current_table_discrete, current_mean_variance_df %>% dplyr::select(c("team", "Mean")))
names(current_table_discrete)[names(current_table_discrete) == 'Mean'] <- 'current_exp_points'

# Discretising current expected points
current_table_discrete$current_exp_points_bin <- cut(current_table_discrete$current_exp_points,
                                 breaks = quantile(current_table_discrete$current_exp_points, probs = c(0, .33, .66, 1), na.rm=TRUE),
                                 include.lowest = TRUE,
                                 labels = c("Low","Med","High"))

# Bin goal differences from end of sesaon tables
season_tables <- discretise_gd(season_tables)

# Converting historical points into ranks
season_tables$rank <- points_to_rank(season_tables$points)

season_tables$elo_bin <- cut(season_tables$elo,
        breaks = quantile(
        season_tables$elo,
        probs = c(0, .25, .5, .75, 1), na.rm=TRUE
      ),
      include.lowest = TRUE,
      labels = c("rating1","rating2","rating3","rating4"))
```

# 5 Functions for fitting the BN

```{r}
# Helper: build empirical marginal or uniform fallback

build_marginal <- function(node_name, levels) {

    # try season_tables first, then current_table_discrete, else uniform
  if (node_name %in% names(season_tables)) {
    tab <- table(as.character(season_tables[[node_name]]))
    vec <- setNames(as.numeric(tab[levels]), names(tab)[match(levels, names(tab))])
    vec[is.na(vec)] <- 0
    if (sum(vec) > 0) return(vec / sum(vec))
  }
  if (node_name %in% names(current_table_discrete)) {
    tab <- table(as.character(current_table_discrete[[node_name]]))
    vec <- setNames(as.numeric(tab[levels]), names(tab)[match(levels, names(tab))])
    vec[is.na(vec)] <- 0
    if (sum(vec) > 0) return(vec / sum(vec))
  }
  # fallback uniform
  setNames(rep(1/length(levels), length(levels)), levels)
}
```

# 5.1 Generating node distributions

```{r}
# -----------------------
# 1. Compute base marginal P(Rank) from historical data
# -----------------------
# season_tables must have column 'rank' (1..14)
n_rank <- 14
base_rank_counts <- season_tables %>%
  count(rank) %>%
  arrange(as.integer(rank))
# ensure all ranks 1..14 present
all_ranks <- as.integer(1:n_rank)
counts_complete <- rep(0, n_rank)
names(counts_complete) <- as.character(all_ranks)
counts_complete[as.character(base_rank_counts$rank)] <- base_rank_counts$n
base_rank_p <- counts_complete / sum(counts_complete)  # marginal P(Rank)

# -----------------------
# 2. Build parent_levels & parent_grid
# -----------------------
# parent levels should come from the union of levels you will use in your BN.
# For hybrid we create a grid of possible parent states to produce P(R|parents)
parent_levels <- list(
  current_exp_points_bin = sort(unique(as.character(current_table_discrete$current_exp_points_bin))),
  goal_difference_bin    = sort(unique(as.character(season_tables$goal_difference_bin))),
  elo_bin                = sort(unique(as.character(season_tables$elo_bin)))
)
parent_grid <- expand.grid(parent_levels, stringsAsFactors = FALSE)

# -----------------------
# 3. Define expert shifts
#    Positive shifts -> improve expected rank (i.e., lower numeric rank)
# -----------------------

# Shift for current expected points node
shift_currexp <- c(Low = -3, Med = 0.0, High = 3.0)

# goal_difference shift (example: larger positive GD -> bigger positive shift)
gd_shift_vals <- c(-3.0, -2, -1.5, -0.5, 0.5, 1.5, 2, 3.0)
names(gd_shift_vals) <- c("(-Inf,-30]", "(-30,-20]", "(-20,-10]", "(-10,0]", "(0,10]", "(10,20]", "(20,30]", "(30,Inf)")
shift_gd <- gd_shift_vals

# ELO rating shift
shift_elo <- c(rating1 = -1, rating2 = 0.0, rating3 = 1, rating4 = 2)

# Ensure shift vectors have proper names consistent with parent_levels
if (!all(names(shift_currexp) %in% parent_levels$current_exp_points_bin)) {
  # attempt to auto-name using parent levels if shift_currexp was built with generic names
  names(shift_currexp) <- parent_levels$current_exp_points_bin[seq_along(shift_currexp)]
}
if (!all(names(shift_gd) %in% parent_levels$goal_difference_bin)) {
  names(shift_gd) <- parent_levels$goal_difference_bin[seq_along(shift_gd)]
}
if (!all(names(shift_elo) %in% parent_levels$elo_bin)) {
  names(shift_elo) <- parent_levels$elo_bin[seq_along(shift_elo)]
}

# -----------------------
# 4. Build CPT matrix using shift rule + smoothing
# -----------------------
cpt_build_res <- build_cpt_from_base_shift3(
  base_rank_p = base_rank_p,
  parent_grid = parent_grid,
  shift_currexp = shift_currexp,
  shift_gd = shift_gd,
  shift_elo = shift_elo,
  sd_multiplier = 1.0,
  alpha = 1.0,
  pseudo_scale = 1000
)
cpt_mat <- cpt_build_res$cpt_mat  # matrix: rows = ranks (1..14), cols = parent configs

# Quick check: columns sum to 1
stopifnot(all(abs(colSums(cpt_mat) - 1) < 1e-8))

# -----------------------
# 5-6. Prepare CPT array and fit BN
# -----------------------

# 1) Convert CPT matrix to array
rank_names <- as.character(seq_len(n_rank))  # 1..14
cpt_arr <- matrix_to_cpt_array3(cpt_mat, parent_levels = parent_levels, rank_names = rank_names)

# 2) Ensure dim_list exists (sync with cpt_arr dimnames)

# Create/restore dim_list from cpt_arr dimnames in a robust way.
if (!exists("dim_list") || is.null(dim_list)) {
  dn <- dimnames(cpt_arr)

  # If per-axis labels are missing, try to build them from parent_levels & rank_names
  if (is.null(dn)) {
    dn <- c(list(rank = rank_names), parent_levels)
  } else {
    # ensure per-axis labels exist (fill missing with parent_levels / rank_names)
    nd <- length(dim(cpt_arr))
    if (length(dn) != nd) {
      # normalize to length
      full_dn <- vector("list", nd)
      full_dn[seq_len(min(length(dn), nd))] <- dn[seq_len(min(length(dn), nd))]
      dn <- full_dn
    }
    # fill missing axis labels using known sources
    for (i in seq_along(dn)) {
      if (is.null(dn[[i]])) {
        # first axis -> rank_names, others -> parent_levels in order
        if (i == 1) {
          dn[[i]] <- as.character(rank_names)
        } else {
          pl_names <- names(parent_levels)
          if (!is.null(pl_names) && length(pl_names) >= (i - 1)) {
            dn[[i]] <- as.character(parent_levels[[pl_names[i - 1]]])
          } else {
            dn[[i]] <- paste0("Level", i, seq_len(dim(cpt_arr)[i]))
          }
        }
      }
    }
  }

  # Ensure axis names (names(dn)) are present and in expected order
  expected_axis_names <- c("rank", names(parent_levels))
  if (length(dn) == length(expected_axis_names)) {
    # if there are no names, set them; if some names present but different length, override safely
    names(dn) <- expected_axis_names
  } else {
    # fallback: set names to HomeRank + parent names for the first matching positions
    names(dn) <- expected_axis_names[seq_along(dn)]
  }

  dim_list <- dn
}

# 3) Handle single-level parents by adding dummy level
# Robust way to add a dummy level for any parent that currently has only one level

dist_list <- list()

for (parent in names(parent_levels)) {
  # find axis position (must exist in dim_list)
  axis_pos <- which(names(dim_list) == parent)
  if (length(axis_pos) == 0) stop("Axis '", parent, "' not found in dim_list/dimnames(cpt_arr).")

  # If the axis currently has only 1 level, expand it to 2 (add "dummy")
  if (length(dim_list[[axis_pos]]) == 1) {
    # current dimnames and dims
    old_dn <- dim_list
    old_names <- names(old_dn)
    dims <- dim(cpt_arr)
    dims[axis_pos] <- 2

    # replicate values of cpt_arr along that axis (repeat each slice)
    # We repeat the flattened values in a way that keeps ordering consistent
    new_vals <- rep(as.vector(cpt_arr), times = 2)
    cpt_arr <- array(new_vals, dim = dims)

    # update dim_list: insert original level then "dummy" at axis_pos
    new_dn <- old_dn
    new_dn[[axis_pos]] <- c(old_dn[[axis_pos]][1], "dummy")
    names(new_dn) <- old_names

    # assign back both dim_list and dimnames of the array
    dim_list <- new_dn
    dimnames(cpt_arr) <- dim_list

    # ensure dist_list has an entry; if not, create a uniform prior for the existing levels
    if (is.null(dist_list[[parent]]) || length(dist_list[[parent]]) == 0) {
      dist_list[[parent]] <- setNames(rep(1 / length(parent_levels[[parent]]),
                                          length(parent_levels[[parent]])),
                                      parent_levels[[parent]])
    }

    # append dummy with prob 0 and renormalize
    dist_list[[parent]] <- c(dist_list[[parent]], dummy = 0)
    if (sum(dist_list[[parent]]) == 0) {
      # no mass — make uniform across the expanded levels
      dist_list[[parent]] <- setNames(rep(1 / length(dist_list[[parent]]),
                                          length(dist_list[[parent]])),
                                      names(dist_list[[parent]]))
    } else {
      dist_list[[parent]] <- dist_list[[parent]] / sum(dist_list[[parent]])
    }

    message(sprintf("Expanded axis '%s' to include dummy. New levels: %s",
                    parent, paste(dim_list[[axis_pos]], collapse = ", ")))
  }
}

# 4) Ensure dimnames are correctly named
dn <- dimnames(cpt_arr)
if (is.null(names(dn))) names(dn) <- c("rank", names(parent_levels))
dimnames(cpt_arr) <- dn

# 5) Reorder array axes to match DAG (child first, then parents)
expected_parents <- names(parent_levels)
perm_order <- match(c("rank", expected_parents), names(dimnames(cpt_arr)))
cpt_arr <- aperm(cpt_arr, perm = perm_order)
```

```{r}
# --- Robust preparation of dist_list and assignment of CPT ---

# Identify child/axis names from cpt_arr
dn <- dimnames(cpt_arr)
axis_names <- names(dn)
if (is.null(axis_names) || length(axis_names) < 1) stop("dimnames(cpt_arr) must have axis names set.")

child_name <- axis_names[1]
expected_parents <- axis_names[-1]

# Ensure dist_list exists
if (!exists("dist_list") || is.null(dist_list)) dist_list <- list()

# For each parent, ensure dist_list[[parent]] is a named prob vector matching dimnames(cpt_arr)[[parent]]
for (parent in expected_parents) {
  desired_levels <- as.character(dimnames(cpt_arr)[[parent]])
  if (is.null(desired_levels) || length(desired_levels) == 0) {
    stop("dimnames(cpt_arr)[[", parent, "]] is NULL/empty. Fix cpt_arr dimnames first.")
  }

  cur <- dist_list[[parent]]

  if (is.null(cur) || length(cur) == 0) {
    # build an empirical prior or uniform fallback
    dist_list[[parent]] <- build_marginal(parent, desired_levels)
    message(sprintf("Initialized dist_list[['%s']] with %d levels (sum=%g).",
                    parent, length(dist_list[[parent]]), sum(dist_list[[parent]])))
  } else {
    # If present but names don't match, align by intersection or fill missing with zeros
    if (is.null(names(cur)) || !all(desired_levels %in% names(cur))) {
      newv <- setNames(numeric(length(desired_levels)), desired_levels)
      if (!is.null(names(cur))) {
        common <- intersect(names(cur), desired_levels)
        if (length(common) > 0) newv[common] <- cur[common]
      }
      if (sum(newv) == 0) {
        # fallback: uniform
        newv[] <- 1 / length(newv)
      } else {
        newv <- newv / sum(newv)
      }
      dist_list[[parent]] <- newv
      message(sprintf("Re-aligned dist_list[['%s']] to desired levels.", parent))
    } else {
      # names match: reorder to desired_levels and normalize
      vec <- as.numeric(cur[desired_levels])
      names(vec) <- desired_levels
      if (sum(vec) == 0) vec[] <- 1 / length(vec)
      dist_list[[parent]] <- vec / sum(vec)
    }
  }
  # Final safety: ensure numeric and sums to 1
  if (!is.numeric(dist_list[[parent]])) stop("dist_list[[", parent, "]] must be numeric.")
  dist_list[[parent]] <- dist_list[[parent]] / sum(dist_list[[parent]])
}

# Assign CPT array to the child node in dist_list
dist_list[[child_name]] <- cpt_arr

# ----------------
# Defining the DAG
# ----------------

model_string <- "[rank|current_exp_points_bin:goal_difference_bin:elo_bin][current_exp_points_bin][goal_difference_bin][elo_bin]"
dag <- model2network(model_string)

# -----------------
# Validation checks
# -----------------

#  1. every node in dag_example must have an entry in dist_list
missing_nodes <- setdiff(nodes(dag), names(dist_list))
if (length(missing_nodes) > 0) stop("dist_list missing entries for nodes: ", paste(missing_nodes, collapse = ", "))

if ("elo_bin" %in% names(dist_list)) {
  lv <- names(dist_list[["elo_bin"]])
  if (any(is.na(lv))) {
    lv[is.na(lv)] <- "dummy"
    names(dist_list[["elo_bin"]]) <- lv
  }
}

#  2. parents' names align
for (parent in expected_parents) {
  if (!all(as.character(dimnames(cpt_arr)[[parent]]) %in% names(dist_list[[parent]]))) {
    stop("Parent levels in cpt_arr and dist_list do not align for parent: ", parent)
  }
}

for (parent in expected_parents) {
  probs <- dist_list[[parent]]
  dist_list[[parent]] <- table(factor(names(probs), levels = names(probs)))  # ensure levels
  dist_list[[parent]][] <- probs  # assign probabilities
}
```

# 5.2 Fitting the BN

```{r}
# Fit BN
fitted_custom <- bnlearn::custom.fit(dag, dist = dist_list)
```

# 6. Generating predictions from the BN

# 6.1 Helper functions for predictions

```{r}
# --------------------------------
# Get rank distribution for a team
# --------------------------------

# sample conditional distribution of `node` given evidence_list
get_rank_dist <- function(fitted_bn, evidence_list, n = 20000, seed = 2025) {
  
  # Seed for reproducibility
  # set.seed(seed)
  
  # Sample rank from the fitted BN based on the values of parent nodes
  samp <- cpdist(fitted_bn, nodes = "rank", evidence = evidence_list, method = "lw", n = n)
  
  levels_all <- if (!is.null(fitted_bn[["rank"]]$prob)) dimnames(fitted_bn[["rank"]]$prob)[[1]] else levels(samp[["rank"]])
  probs <- prop.table(table(factor(samp[["rank"]], levels = levels_all)))
  setNames(as.numeric(probs), names(probs))
}

# ---------------------------------
# Get parent node values for a team
# ---------------------------------

build_evidence <- function(row, prefix = "home_", fitted_bn) {
  
  # Parent nodes of rank node
  parents <- fitted_bn[["rank"]]$parents
  
  vals <- lapply(parents, function(p) {
    col <- paste0(prefix, p)
    as.character(row[[col]])    # scalar, with factor label (not internal code)
  })
  names(vals) <- parents
  vals   # named list of length-1 character scalars
}

# --------------------
# Predict match result
# --------------------

predict_match <- function(fitted_bn, match_cpt, home_ev, away_ev, n = 20000, seed = 2025) {
  
  # Get rank distribution for home team
  h <- get_rank_dist(fitted_bn, home_ev, n = n, seed = seed)
  
  # Get rank distribution for away team
  a <- get_rank_dist(fitted_bn, away_ev, n = n, seed = seed + 1)
  
  # Compute the joint probability of each rank pair
  # P(rank_h, rank_a) = P(rank_h | evidence_h)*P(rank_a | evidence_a)
  P_pair <- h[as.character(match_cpt$rank_home)] * a[as.character(match_cpt$rank_away)]
    
  # P(H) = P(rank_h, rank_a)*P(H | rank_h, rank_a) summed over all rank pairs
  p_H <- sum(P_pair * match_cpt$p_H)
  # P(D) = P(rank_h, rank_a)*P(D | rank_h, rank_a) summed over all rank pairs
  p_D <- sum(P_pair * match_cpt$p_D)
  # P(A) = P(rank_h, rank_a)*P(A | rank_h, rank_a) summed over all rank pairs
  p_A <- sum(P_pair * match_cpt$p_A)
  
  probs <- c(p_H = p_H, p_D = p_D, p_A = p_A)
  # Normalise probabilities to 1
  probs / sum(probs)
}

# -------------------------------------
# Function to predict multiple fixtures
# -------------------------------------

predict_fixtures <- function(fixtures_df, fitted_bn, match_cpt, n = 20000, seed = 2025) {
  
  preds <- t(sapply(seq_len(nrow(fixtures_df)), function(i) {
    row <- fixtures_df[i, , drop = FALSE]
    home_ev <- build_evidence(row, "home_", fitted_bn)
    away_ev <- build_evidence(row, "away_", fitted_bn)
    predict_match(fitted_bn, match_cpt, home_ev, away_ev, n = n, seed = seed + i)
  }))
  cbind(fixtures_df, as.data.frame(preds))
}

# ------------------------------------------------
# Function to generate tables after each gameweek
# ------------------------------------------------

build_standings_by_gw <- function(results2,
                                  matches_per_gameweek = 10,
                                  home_col = "home_team", away_col = "away_team",
                                  hg_col = "home_goals", ag_col = "away_goals",
                                  result_col = "result", match_id_col = "match_id",
                                  home_elo_col = "home_elo", away_elo_col = "away_elo") {
  
  # turn each match into two rows (one per team) and include elo
  long <- bind_rows(
    results2 %>%
      transmute(match_id, gameweek,
                team = .[[home_col]],
                gf = .[[hg_col]],
                ga = .[[ag_col]],
                result = .[[result_col]],
                elo = .[[home_elo_col]]) %>%
      mutate(points = case_when(result == "H" ~ 3, result == "D" ~ 1, TRUE ~ 0)),
    results2 %>%
      transmute(match_id, gameweek,
                team = .[[away_col]],
                gf = .[[ag_col]],
                ga = .[[hg_col]],
                result = .[[result_col]],
                elo = .[[away_elo_col]]) %>%
      mutate(points = case_when(result == "A" ~ 3, result == "D" ~ 1, TRUE ~ 0))
  ) %>%
    arrange(team, gameweek, match_id)

  # cumulative sums per team across gameweeks (by match order)
  team_gw_cum <- long %>%
    group_by(team) %>%
    arrange(gameweek, match_id) %>%
    mutate(
      gd = gf - ga,
      cum_points = cumsum(points),
      cum_gd = cumsum(gd),
      cum_gf = cumsum(gf),
      last_elo = elo   # store last known elo per match
    ) %>%
    group_by(team, gameweek) %>%
    slice_tail(n = 1) %>%
    ungroup() %>%
    dplyr::select(team, gameweek,
           points = cum_points,
           goal_difference = cum_gd,
           goals_for = cum_gf,
           elo = last_elo)

  # full grid: every team × every gameweek (so teams that didn't play retain last known totals)
  all_gw <- seq_len(max(results2$gameweek, na.rm = TRUE))
  teams <- sort(unique(long$team))
  full <- expand_grid(team = teams, gameweek = all_gw) %>%
    left_join(team_gw_cum, by = c("team", "gameweek")) %>%
    group_by(team) %>%
    arrange(gameweek) %>%
    # carry forward last totals for weeks with no match
    mutate(points = zoo::na.locf(points, na.rm = FALSE),
           goal_difference = zoo::na.locf(goal_difference, na.rm = FALSE),
           goals_for = zoo::na.locf(goals_for, na.rm = FALSE),
           elo = zoo::na.locf(elo, na.rm = FALSE)) %>%
    replace_na(list(points = 0, goal_difference = 0, goals_for = 0, elo = NA)) %>%
    ungroup()

  # produce final tables per gameweek and compute position by tie-breakers
  standings_long <- full %>%
    group_by(gameweek) %>%
    arrange(gameweek,
            desc(points),
            desc(goal_difference),
            desc(goals_for),
            team) %>%                # stable tie-breaker: alphabetical
    mutate(position = row_number()) %>%
    ungroup() %>%
    dplyr::select(gameweek, position, team, points, goal_difference, elo) %>%
    arrange(gameweek, position)

  standings_long
}

# ----------------------------------------
# Function to add expected points to table
# ----------------------------------------

add_exp_points_to_standings <- function(results, standings_by_gw, season_points,
                                        compute_component1_mean,
                                        cur_season, matches_per_gameweek = 10) {
  # ensure results have a gameweek column
  results2 <- if ("gameweek" %in% names(results)) {
    results
  } else {
    print("no")
    results %>% mutate(gameweek = ceiling(as.numeric(match_id) / matches_per_gameweek))
  }

  gws <- sort(unique(standings_by_gw$gameweek))

  comp_by_gw <- lapply(gws, function(gw) {
    # 1) current points from standings (points after this gw)
    cur_pts <- standings_by_gw %>%
      filter(gameweek == gw) %>%
      dplyr::select(team, total_points = points, elo)  # include elo here

    # 2) matches played up to and including this gw
    mp <- results2 %>%
      filter(gameweek <= gw) %>%
      transmute(team_home = home_team, team_away = away_team) %>%
      # stack home/away to count matches per team
      { bind_rows(
          transmute(., team = team_home),
          transmute(., team = team_away)
        ) } %>%
      count(team, name = "matches_played")

    # 3) ensure every team in standings appears, missing -> 0 matches
    cur_season_points <- cur_pts %>%
      left_join(mp, by = "team") %>%
      mutate(matches_played = ifelse(is.na(matches_played), 0L, matches_played))

    # 4) call your function to compute component1 (returns tibble team, Mean, Var)
    comp <- compute_component1_mean(season_points, cur_season_points, cur_season)

    # attach gameweek
    comp %>% mutate(gameweek = gw) %>% dplyr::select(gameweek, team, Mean, Var)
  })

  comp_df <- bind_rows(comp_by_gw)

  # 5) join Mean (and Var) back to standings_by_gw
  standings_by_gw %>%
    left_join(comp_df, by = c("gameweek", "team")) %>%
    arrange(gameweek, position) %>%
    # ensure elo column is preserved
    dplyr::select(gameweek, position, team, points, goal_difference, elo, everything())
}

```

# Function to generate predictions for a full season

```{r, warning=FALSE}
get_predictions_for_full_season <- function(
    season_int, historical_matches, season_tables, fitted_bn, cpt, results2526, current_gw) {
    
  matches_per_gameweek <- 10
  
  if (season_int == 2025) {
    season_results <- results2526 %>%
      rename(match_id = match_number) %>%
      filter(gameweek <= current_gw)
    season_results$season <- 2025
  } else {
  
    season_results <- historical_matches[historical_matches$season == season_int, ]
    season_results$match_id <- season_results$match_id - min(season_results$match_id) + 1
    
    season_results <- season_results %>%
    mutate(
      match_id = .[["match_id"]],
      gameweek = ceiling(as.numeric(match_id) / matches_per_gameweek)
    )
  }
  
  standings_by_gameweek <- build_standings_by_gw(season_results, matches_per_gameweek = matches_per_gameweek)
  
  standings_with_expected_points <- add_exp_points_to_standings(
    results = season_results,
    standings_by_gw = standings_by_gameweek,
    season_points = season_tables,
    compute_component1_mean = compute_component1_mean,
    cur_season = season_int + 1,
    matches_per_gameweek = matches_per_gameweek
  )
  
  # Calculate projected goal difference
  standings_with_expected_points$projected_goal_difference <- 
    standings_with_expected_points$goal_difference / 
    standings_with_expected_points$gameweek * 38
  
  # Rounding
  standings_with_expected_points <- standings_with_expected_points %>% 
    mutate(across("projected_goal_difference", round, 0))
  
  names(standings_with_expected_points)[names(standings_with_expected_points) == 'Mean'] <- 'current_exp_points'
  names(standings_with_expected_points)[names(standings_with_expected_points) == 'elo'] <- 'rating'
  
  standings_with_expected_points <- discretise_features(
    df = standings_with_expected_points
  )
  
  standings_with_expected_points$current_exp_points_bin <- 
    cut(standings_with_expected_points$current_exp_points,
        breaks = quantile(
          standings_with_expected_points$current_exp_points,
          probs = c(0, .33, .66, 1), na.rm=TRUE
        ),
        include.lowest = TRUE,
        labels = c("Low","Med","High"))
  
  # Merging on the node columns (expected points, goal difference, ELO) with fixtures
  # Merge on current points, goal difference and elo
  season_results <- season_results %>%
    left_join(
      standings_with_expected_points %>%
        dplyr::select("gameweek", "team", "goal_difference_bin", "elo_bin", "current_exp_points_bin") %>%
        rename(
          home_team                   = team,
          home_goal_difference_bin    = goal_difference_bin,
          home_elo_bin                = elo_bin,
          home_current_exp_points_bin = current_exp_points_bin
        ),
      by = c("home_team", "gameweek")
    ) %>%
    left_join(
      standings_with_expected_points %>%
        dplyr::select("gameweek", "team", "goal_difference_bin", "elo_bin", "current_exp_points_bin") %>%
        rename(
          away_team                   = team,
          away_goal_difference_bin    = goal_difference_bin,
          away_elo_bin                = elo_bin,
          away_current_exp_points_bin = current_exp_points_bin
        ),
      by = c("away_team", "gameweek")
    )
  
    # Generate predictions
    predictions <- predict_fixtures(
      fixtures_df = season_results,
      fitted_bn = fitted_bn,
      match_cpt = cpt
    )
    
    # Sampling home and away ranks
    predictions <- predictions %>%
      rowwise() %>%
      mutate(
        sim_rank_home = {
          rd <- get_rank_dist(
            fitted_bn,
            evidence_list = list(
              elo_bin                = home_elo_bin,
              goal_difference_bin    = home_goal_difference_bin,
              current_exp_points_bin = home_current_exp_points_bin
            )
          )
          as.integer(sample(names(rd), size = 1, prob = rd))
        },
        
        sim_rank_away = {
          rd <- get_rank_dist(
            fitted_bn,
            evidence_list = list(
              elo_bin                = away_elo_bin,
              goal_difference_bin    = away_goal_difference_bin,
              current_exp_points_bin = away_current_exp_points_bin
            )
          )
          as.integer(sample(names(rd), size = 1, prob = rd))
        }
      ) %>%
      ungroup()
    
    return(predictions)
}
```

# 6.2  Building the CPT for home and away rank

```{r}
# Building CPT for home and away rank combinations
tmp <- build_empirical_cpt(
  historical_matches = historical_matches,
  season_points = season_tables,
  lambda = 1
)

cpt_array <- tmp$cpt_array
cpt_wide <- tmp$cpt_wide
```

# 6.3 Filtering fixtures for one gameweek and adding node values

```{r}
# Filter for current week fixtures
current_fixtures <- fixtures_2526[fixtures_2526$gameweek == 8, ]

# Merge on current points, goal difference and elo
current_fixtures <- current_fixtures %>%
  left_join(
    current_table_discrete %>%
      dplyr::select("team", "goal_difference_bin", "elo_bin", "current_exp_points_bin") %>%
      rename(
        home_team                   = team,
        home_goal_difference_bin    = goal_difference_bin,
        home_elo_bin                = elo_bin,
        home_current_exp_points_bin = current_exp_points_bin
      ),
    by = "home_team"
  ) %>%
  left_join(
    current_table_discrete %>%
      dplyr::select("team", "goal_difference_bin", "elo_bin", "current_exp_points_bin") %>%
      rename(
        away_team                   = team,
        away_goal_difference_bin    = goal_difference_bin,
        away_elo_bin                = elo_bin,
        away_current_exp_points_bin = current_exp_points_bin
      ),
    by = "away_team"
  )
```

# 6.4 Generating predictions

# Generating predictions for the current gameweek (2025/26)

```{r}
# Generate predictions
current_gameweek_predictions <- predict_fixtures(current_fixtures, fitted_custom, cpt_wide)
```

# Generatings predictions for each season since 2000/01

```{r, warning=FALSE}
all_predictions <- list()

for (season_int in (2000:2024)){
  message(glue("Generating predictions for season {season_int}/{season_int + 1}"))
  
  all_predictions[[as.character(season_int)]] <- 
    get_predictions_for_full_season(
      season_int = season_int,
      historical_matches = historical_matches,
      season_tables = season_tables,
      fitted_bn = fitted_custom,
      cpt = cpt_wide,
      results2526 = results_2526,
      current_gw = 28
    )
}
```

### Sampling results and scorelines

```{r}
# Get historical matches with EOS ranks added
historical_matches_with_rank <- tmp$ranked_matches

score_dist <- historical_matches_with_rank %>%
  mutate(scoreline = paste0(home_goals, "-", away_goals)) %>%
  group_by(rank_home, rank_away, result, scoreline) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(rank_home, rank_away, result) %>%
  mutate(prob = n / sum(n)) %>%
  ungroup()

# Dirichlet smoothing
score_dist_smooth <- historical_matches_with_rank %>%
  mutate(scoreline = paste0(home_goals, "-", away_goals)) %>%
  group_by(rank_home, rank_away, result, scoreline) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(rank_home, rank_away, result) %>%
  tidyr::complete(scoreline, fill = list(n = 0)) %>%
  mutate(total = sum(n),
         prob = (n + 1) / (total + n_distinct(scoreline))) %>%
  ungroup()

# Sample the scoreline based on learned CPT
sample_score <- function(home_rank, away_rank, outcome, score_dist) {
  
  home_rank <- as.numeric(home_rank)
  away_rank <- as.numeric(away_rank)
  
  subset <- score_dist %>%
    filter(
      rank_home == home_rank,
      rank_away == away_rank,
      result == outcome
    )
  
  if (nrow(subset) == 0) {
    
    if (outcome == "H") {
      return("1-0")  # temporary fallback 
    }
    else if (outcome == "D") {
      return("1-1")  # temporary fallback 
    }
    else if (outcome == "A") {
      return("0-1")  # temporary fallback 
    }
  }
  
  if (any(is.na(subset$prob)) || sum(subset$prob) == 0) {

    print(subset)
    return("1-0")
  }
  
  sample(subset$scoreline, size = 1, prob = subset$prob)
}

# Sample result from predictions
sample_result <- function(p_H, p_D, p_A) {
  sample(c("H","D","A"), size = 1, prob = c(p_H, p_D, p_A))
}
```

# Sampling result and scoreline based on predictions

```{r}
for (season_int in (2024:2024)){
  message(glue("Sampling scorelines for season: {season_int}/{season_int+1}"))
  
  all_predictions[[as.character(season_int)]] <- 
    all_predictions[[as.character(season_int)]] %>%
    rowwise() %>%
    mutate(
      sim_result = sample_result(p_H, p_D, p_A),
      sim_score  = sample_score(sim_rank_home, sim_rank_away, sim_result, score_dist),
      sim_home_goals = as.numeric(sub("-.*", "", sim_score)),
      sim_away_goals = as.numeric(sub(".*-", "", sim_score))
    ) %>%
    ungroup()
}
```
# Simulating seasons

```{r}
full_season_simulation <- function(
  season_int,
  fixtures,
  historical_matches,
  season_tables,
  fitted_bn,
  cpt,
  score_dist,
  initial_elo,
  k_factor = 20
) {
  
  matches_per_gameweek <- 10
  
  # --- INITIAL STATE ---
  
  teams <- unique(c(fixtures$home_team, fixtures$away_team))
  
  current_table <- data.frame(
    team = teams,
    points = 0,
    goals_for = 0,
    goals_against = 0,
    goal_difference = 0
  )
  
  current_elo <- initial_elo
  current_elo <- setNames(initial_elo$elo, initial_elo$team)

  simulated_results <- data.frame(
    home_team   = character(),
    away_team   = character(),
    gameweek    = integer(),
    home_goals  = numeric(),
    away_goals  = numeric(),
    sim_result  = character(),
    stringsAsFactors = FALSE
  )
  
  # --- LOOP OVER GAMEWEEKS ---
  
  for (gw in sort(unique(fixtures$gameweek))) {
    
    gw_matches <- fixtures[fixtures$gameweek == gw, ]
    
    # --- BUILD STANDINGS ---
    
    current_table <- current_table %>%
      mutate(
        gameweek = gw,
        projected_goal_difference = round(goal_difference / gw * 38, 0)
      )
    
    # --- EXPECTED POINTS ---
    
    # Add ELO
    current_table$elo <- current_elo[current_table$team]
    
    # Add gameweek
    current_table$gameweek <- gw
    
    # Add position
    current_table <- current_table %>%
      arrange(desc(points), desc(goal_difference)) %>%
      mutate(position = row_number())
    
    standings_with_exp <- add_exp_points_to_standings(
      results = simulated_results,
      standings_by_gw = current_table,
      season_points = season_tables,
      compute_component1_mean = compute_component1_mean,
      cur_season = season_int,
      matches_per_gameweek = matches_per_gameweek
    )
    
    names(standings_with_exp)[names(standings_with_exp) == "Mean"] <- "current_exp_points"
    
    # --- DISCRETISE FEATURES ---
    
    standings_with_exp <- discretise_features(standings_with_exp)
    
    standings_with_exp$current_exp_points_bin <- cut(
      standings_with_exp$current_exp_points,
      breaks = quantile(
        standings_with_exp$current_exp_points,
        probs = c(0, .33, .66, 1),
        na.rm = TRUE
      ),
      include.lowest = TRUE,
      labels = c("Low", "Med", "High")
    )
    
    # --- MERGE FEATURES INTO MATCHES ---
    
    gw_matches <- gw_matches %>%
      left_join(
        standings_with_exp %>%
          dplyr::select(team, goal_difference_bin, elo_bin, current_exp_points_bin) %>%
          rename(
            home_team = team,
            home_goal_difference_bin = goal_difference_bin,
            home_elo_bin = elo_bin,
            home_current_exp_points_bin = current_exp_points_bin
          ),
        by = "home_team"
      ) %>%
      left_join(
        standings_with_exp %>%
          dplyr::select(team, goal_difference_bin, elo_bin, current_exp_points_bin) %>%
          rename(
            away_team = team,
            away_goal_difference_bin = goal_difference_bin,
            away_elo_bin = elo_bin,
            away_current_exp_points_bin = current_exp_points_bin
          ),
        by = "away_team"
      )
    
    # --- PREDICT ---
    
    gw_preds <- predict_fixtures(
      fixtures_df = gw_matches,
      fitted_bn = fitted_bn,
      match_cpt = cpt
    )
    
    # --- SAMPLE RANKS ---
    
    gw_preds <- gw_preds %>%
      rowwise() %>%
      mutate(
        sim_rank_home = {
          rd <- get_rank_dist(
            fitted_bn,
            evidence_list = list(
              elo_bin = home_elo_bin,
              goal_difference_bin = home_goal_difference_bin,
              current_exp_points_bin = home_current_exp_points_bin
            )
          )
          as.integer(sample(names(rd), 1, prob = rd))
        },
        sim_rank_away = {
          rd <- get_rank_dist(
            fitted_bn,
            evidence_list = list(
              elo_bin = away_elo_bin,
              goal_difference_bin = away_goal_difference_bin,
              current_exp_points_bin = away_current_exp_points_bin
            )
          )
          as.integer(sample(names(rd), 1, prob = rd))
        }
      ) %>%
      ungroup()
    
    # --- SIMULATE SCORES ---
    
    gw_preds <- gw_preds %>%
      rowwise() %>%
      mutate(
        sim_result = sample_result(p_H, p_D, p_A),
        sim_score  = sample_score(sim_rank_home, sim_rank_away, sim_result, score_dist),
        home_goals = as.numeric(sub("-.*", "", sim_score)),
        away_goals = as.numeric(sub(".*-", "", sim_score))
      ) %>%
      ungroup()
    
    # --- UPDATE TABLE ---
    
    for (i in 1:nrow(gw_preds)) {
      
      h <- gw_preds$home_team[i]
      a <- gw_preds$away_team[i]
      
      hg <- gw_preds$home_goals[i]
      ag <- gw_preds$away_goals[i]
      
      # goals
      current_table$goals_for[current_table$team == h] <- current_table$goals_for[current_table$team == h] + hg
      current_table$goals_against[current_table$team == h] <- current_table$goals_against[current_table$team == h] + ag
      
      current_table$goals_for[current_table$team == a] <- current_table$goals_for[current_table$team == a] + ag
      current_table$goals_against[current_table$team == a] <- current_table$goals_against[current_table$team == a] + hg
      
      # points
      if (hg > ag) {
        current_table$points[current_table$team == h] <- current_table$points[current_table$team == h] + 3
      } else if (hg < ag) {
        current_table$points[current_table$team == a] <- current_table$points[current_table$team == a] + 3
      } else {
        current_table$points[current_table$team == h] <- current_table$points[current_table$team == h] + 1
        current_table$points[current_table$team == a] <- current_table$points[current_table$team == a] + 1
      }
      
      # --- UPDATE ELO ---
      
      home_advantage <- 60
      
      expected_home <- 1 / (1 + 10 ^ ((current_elo[a] - (current_elo[h] + home_advantage)) / 400))
      
      actual_home <- ifelse(hg > ag, 1, ifelse(hg == ag, 0.5, 0))
      
      new_home <- current_elo[h] + k_factor * (actual_home - expected_home)
      new_away <- current_elo[a] + k_factor * ((1 - actual_home) - (1 - expected_home))
      
      current_elo[h] <- new_home
      current_elo[a] <- new_away
    }
    
    # recompute GD
    current_table$goal_difference <- current_table$goals_for - current_table$goals_against
    
    # store results
    simulated_results <- rbind(simulated_results, gw_preds)
  }
  
  # --- FINAL TABLE ---
  
  final_table <- current_table %>%
    arrange(desc(points), desc(goal_difference))
  
  final_table$position <- seq_len(nrow(final_table))
  
  return(final_table)
}
```

# Multiple simulations for 2024

```{r}
set.seed(12345)
n_sims <- 500

sim_tables <- lapply(1:n_sims, function(i) {
  
  message(glue("Running simulation ", i, " for season {season_int}/{season_int + 1}"))
  
  tmp_sim <- full_season_simulation(
    season_int = 2024,
    historical_matches = historical_matches,
    season_tables = season_tables,
    fitted_bn = fitted_custom,
    cpt = cpt_wide,
    score_dist = score_dist,
    initial_elo = initial_elo_24,
    fixtures = fixtures_2024,
    k_factor = 20
  )
})

# Combine simulations
all_results <- do.call(rbind, 
  lapply(seq_along(sim_tables), function(i) {
    table <- sim_tables[[i]]
    table$simulation <- i
    table
  })
)

final_simulation <- aggregate(
  cbind(points, goal_difference, position) ~ team,
  data = all_results,
  FUN = mean
)

colnames(final_simulation) <- c(
  "team",
  "avg_points",
  "avg_gd",
  "avg_position"
)

# Sort by expected strength
final_simulation <- final_simulation[
  order(-final_simulation$avg_points, -final_simulation$avg_gd),
]

# Add expected league position
final_simulation$position <- seq_len(nrow(final_simulation))

rownames(final_simulation) <- NULL

# Merge on true table
simulation_vs_actual <- final_simulation %>%
  rename(sim_position = position) %>%
  left_join(season_tables[season_tables$season == 2024,] 
            %>% dplyr::select(team,position) 
            %>% rename(actual_position = position), 
            by = c("team"))
```

# Champions, top 4, and relegation probabilities under the simulations

```{r}
sim_probabilities <- all_results %>%
  group_by(team) %>%
  summarise(
    P_title = mean(position == 1),
    P_top4 = mean(position <= 4),
    P_relegation = mean(position >= 18),
    avg_position = mean(position)
  ) %>%
  arrange(avg_position)
```

# 7. Testing / validating predictions

# Getting current season results

```{r}
# Merge current results with predictions

current_gameweek_predictions <- current_gameweek_predictions %>%
  left_join(results_2526 %>% dplyr::select(home_team, away_team, result),
            by = c("home_team", "away_team"))

current_season_predictions <- all_predictions$`2024` %>%
  left_join(results_2526 %>% dplyr::select(home_team, away_team, result),
            by = c("home_team", "away_team"))
```

```{r}
# Extract predicted probabilities
pmat_current <- as.matrix(current_gameweek_predictions[, c("p_H", "p_D", "p_A")])

# Convert observed factor/char to ordered integers
obs_vec_current <- match(current_gameweek_predictions$result, c("H", "D", "A"))

rps_values_current <- rps(obs = obs_vec_current, pred = pmat_current)
```

# Pi-Football predictions

```{r}
# -------------------
# Pi football 2010/11
# -------------------

# Loading in predictions from the pi-football model
pi_foot_predictions1011 <- read.csv("Data/pifootball_predictions_1011.csv")

pi_foot_predictions1011[,c("p_H", "p_D","p_A")] <- pi_foot_predictions1011[,c("p_H", "p_D","p_A")] / 100

# Extract predicted probabilities
pmat_pifoot1011 <- as.matrix(pi_foot_predictions1011[, c("p_H", "p_D", "p_A")])

# Convert observed factor/char to ordered integers
obs_vec_pifoot1011 <- match(pi_foot_predictions1011$result, c("H", "D", "A"))

rps_values_pifoot1011 <- rps(obs = obs_vec_pifoot1011, pred = pmat_pifoot1011)


# -------------------
# Pi football 2011/12
# -------------------

# Loading in predictions from the pi-football model
pi_foot_predictions1112 <- read.csv("Data/pifootball_predictions_1112.csv")

pi_foot_predictions1112[,c("p_H", "p_D","p_A")] <- pi_foot_predictions1112[,c("p_H", "p_D","p_A")] / 100

# Extract predicted probabilities
pmat_pifoot1112 <- as.matrix(pi_foot_predictions1112[, c("p_H", "p_D", "p_A")])

# Convert observed factor/char to ordered integers
obs_vec_pifoot1112 <- match(pi_foot_predictions1112$result, c("H", "D", "A"))

rps_values_pifoot1112 <- rps(obs = obs_vec_pifoot1112, pred = pmat_pifoot1112)
```

```{r}
rps_values_pifoot1011$rps
rps_values_pifoot1112$rps
```

# 7.2 Calculating RPS for each season

```{r}
# Generating RPS values for each season

pmat_list <- list()
obs_vec_list <- list()
rps_values_list <- list()

for (season_int in 2000:2024){
  
  pmat_list[[as.character(season_int)]] <- as.matrix(
    all_predictions[[as.character(season_int)]][, c("p_H", "p_D", "p_A")]
  )
  
  obs_vec_list[[as.character(season_int)]] <- match(
    all_predictions[[as.character(season_int)]]$result, c("H", "D", "A")
  )
  
  rps_values_list[[as.character(season_int)]] <- rps(
    obs = obs_vec_list[[as.character(season_int)]], pred = pmat_list[[as.character(season_int)]]
  )$rps

}

# Converting into dataframe
rps_values <- data.frame(
  season = names(rps_values_list),
  rps     = unlist(rps_values_list)
)

# Adding reference row for pi football value
rps_values[nrow(rps_values) + 1,] = c("2010_pifootball", rps_values_pifoot1011$rps)
rps_values[nrow(rps_values) + 1,] = c("2011_pifootball", rps_values_pifoot1112$rps)
```

# Cumulative gameweek comparison with pi-football

```{r, warning=FALSE}
rps_calc <- function(obs_i, pred_row) {
  verification::rps(obs = obs_i, pred = matrix(pred_row, nrow = 1))$rps
}

pmat_1011 <- pmat_list[["2010"]]
obs_vec_1011 <- obs_vec_list[["2010"]]
predictions1011 <- all_predictions[["2010"]]

pmat_1112 <- pmat_list[["2011"]]
obs_vec_1112 <- obs_vec_list[["2011"]]
predictions1112 <- all_predictions[["2011"]]

temp_rps_A <- sapply(seq_len(nrow(pmat_1011)), function(i) rps_calc(obs_vec_1011[i], pmat_1011[i, ]))
temp_rps_B <- sapply(seq_len(nrow(pmat_pifoot1011)), function(i) rps_calc(obs_vec_pifoot1011[i], pmat_pifoot1011[i, ]))

rps_df_1011 <- data.frame(
  match_id = predictions1011$match_id,
  date = predictions1011$date,
  home = predictions1011$home_team, 
  away = predictions1011$away_team,
  rps_challenger = temp_rps_A,
  rps_pifoot = temp_rps_B
) %>%
  mutate(
    rps_diff = temp_rps_B - temp_rps_A,      # positive => B worse than A
    cum_diff = cumsum(rps_diff)    # cumulative difference across matches
  )

temp_rps_A <- sapply(seq_len(nrow(pmat_1112)), function(i) rps_calc(obs_vec_1112[i], pmat_1112[i, ]))
temp_rps_B <- sapply(seq_len(nrow(pmat_pifoot1112)), function(i) rps_calc(obs_vec_pifoot1112[i], pmat_pifoot1112[i, ]))

rps_df_1112 <- data.frame(
  match_id = predictions1112$match_id,
  date = predictions1112$date,
  home = predictions1112$home_team, 
  away = predictions1112$away_team,
  rps_challenger = temp_rps_A,
  rps_pifoot = temp_rps_B
) %>%
  mutate(
    rps_diff = temp_rps_B - temp_rps_A,      # positive -> B worse than A
    cum_diff = cumsum(rps_diff)    # cumulative difference across matches
  )

rps_df_1011$season <- "2010/11"
rps_df_1112$season <- "2011/12"

rps_long_combined <- rbind(rps_df_1011, rps_df_1112)

p_combined_pifoot <- ggplot(rps_long_combined, aes(x = match_id, y = cum_diff)) +
  geom_line(size = 1) +
  geom_point(size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~season, ncol = 2, scales = "fixed") + 
  labs(
    title = "Cumulative Difference in Per-Match RPS (Pi-Football − Challenger)",
    subtitle = "Positive values indicate the proposed model outperforms Pi-Football",
    x = "Match ID (1–380)",
    y = "Cumulative RPS Difference"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold")
  )

p_combined_pifoot
```

# Plotting the average RPS values across seasons

```{r}
rps_long_filtered <- rps_long %>%
  dplyr::filter(model != "rps_SL")

means_df_filtered <- means_df %>%
  dplyr::filter(model != "rps_SL")

rps_long_run_plot <- ggplot(rps_long_filtered[1:25,], 
       aes(x = season, y = rps, color = model, group = model)) +
  geom_line() +
  geom_point() +

  geom_hline(data = means_df_filtered,
             aes(yintercept = mean_rps, color = model),
             linetype = "dashed") +
  
  geom_point(
    data = data.frame(
      season = c("2010", "2011"),
      rps = c(
        rps_long[rps_long$season == "2010_pifootball",]$rps[1],
        rps_long[rps_long$season == "2011_pifootball",]$rps[1]
      ),
      model = "Pi Football"
    ),
    aes(x = season, y = rps, color = model),
    size = 2
  ) +

  labs(
    title = "RPS Values Across Seasons 2000/01 - 2024/25",
    x = "Season Start Year",
    y = "RPS"
  ) +
  scale_color_manual(
    values = c(
      "rps" = "#F8766D",  
      "Pi Football" = "#00BFC4"
    ),
    labels = c(
      "rps" = "Expert Model",
      "Pi Football" = "Pi Football"
    )
  ) +
  labs(colour = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

rps_long_run_plot
```


## Full DAG for completeness

```{r}
# Final BN

full_network_string <- paste0(
  # Home side latent performance factors
  "[Current_Exp_Points_home][GoalDiff_home][Elo_home][RecentForm_home][DaysRest_home]",
  "[HomeRank|Current_Exp_Points_home:GoalDiff_home:Elo_home]",
  "[ExpectedPerformance_home|HomeRank:DaysRest_home:RecentForm_home]",
  
  # Away side latent performance factors
  "[Current_Exp_Points_away][GoalDiff_away][Elo_away][RecentForm_away][DaysRest_away]",
  "[AwayRank|Current_Exp_Points_away:GoalDiff_away:Elo_away]",
  "[ExpectedPerformance_away|AwayRank:DaysRest_away:RecentForm_away]",
  
  # Match-level contextual variable
  "[HomeAdvantage]",
  
  # Outcome
  "[MatchResult|ExpectedPerformance_home:ExpectedPerformance_away:HomeAdvantage]"
)

full_dag <- model2network(full_network_string)

graphviz.plot(full_dag)

full_dag <- model2network(full_network_string)

graphviz.plot(full_dag)
```

## Variance curves for appendix

```{r}
# Plot 1: Previous seasons variance
p1 <- ggplot(df_prev, aes(x = season, y = variance)) +
  geom_line(size = 1) +
  labs(
    x = "Preceding season",
    y = "Variance"
  ) +
  theme_minimal()

# Plot 2: Current season variance
p2 <- ggplot(df_cur, aes(x = matches, y = variance)) +
  geom_line(size = 1) +
  labs(
    x = "Gameweek",
    y = "Variance"
  ) +
  theme_minimal()

# Combine
p1 + p2

ggsave("variance_curves.png", width = 12, height = 4)
```

